### 2021-10-14 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_ncbitaxonomy_softwarenote@bdd677dbefd212c196af0b3f347485d670f31ef2 🚀

\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (\textbf{Federhen2012NcbTax?};
\textbf{Schoch2020NcbTax?}); The basic search functions are coupled with
quality-of-life functions including case-insensitive search and custom
fuzzy string matching to facilitate the amount of information that can
be extracted automatically while allowing efficient manual curation and
inspection of results. \texttt{NCBITaxonomy.jl} works with version 1.6
of the Julia programming language (\textbf{Bezanson2017JulFre?}), and
relies on the Apache Arrow format to store a local copy of the NCBI raw
taxonomy files. The design of \texttt{NCBITaxonomy.jl} has been inspired
by similar efforts, like the R package \texttt{taxadb}
(\textbf{Norman2020TaxHig?}), which provides an offline alternative to
packages like \texttt{taxize} (\textbf{Chamberlain2013TaxTax?}).

@@

Unambiguously identifying species is a far more challenging task than it
may appear. There are a vast number of reasons for this. Different
databases keep different taxonomic ``backbones,'' \emph{i.e.} different
data structures in which names are mapped to species, and organised in a
hierarchy. Not all names are unique identifiers to groups. For example,
\emph{Io} can either refer to a genus of plants from the aster family,
or to a genus of molluscs; the genus \emph{Mus} (of which the house
mouse \emph{Mus musculus} is a species), contains a sub-genus
\emph{also} named \emph{Mus}. Conversely, the same species can have
several names, which are valid synonyms: for example, the domestic cow
\emph{Bos taurus} admits \emph{Bos primigenius taurus} as a valid
synonym. Taxonomic nomenclature also changes regularly, with groups
being split, merged, or moved to a new position in the tree of life;
this is, notably, a common occurrence with viral taxonomy, each
subsequent version of which can differ markedly from the last; compare,
\emph{e.g} (\textbf{Lefkowitz2018VirTax?}) to
(\textbf{Walker2020ChaVir?}).

To add to the complexity, one must also consider that most taxa names
are at some point manually typed, which has the potential to introduce
additional mistakes in raw data; it is likely to expect that such
mistakes may arise when attempting to write down the (perfectly valid)
names of the bacterial isolate known as \emph{Myxococcus
llanfairpwllgwyngyllgogerychwyrndrobwllllantysiliogogogochensis}, or of
the crowned slaty flycatcher \emph{Griseotyrannus
aurantioatrocristatus}. These mistakes are more likely when dealing with
hyper-diverse samples, like plant census (\textbf{Dauncey2016ComMis?};
\textbf{Wagner2016RevSof?}; \textbf{Conti2021MatAlg?}). In addition to
binomial names, the same species can be known by many vernacular
(common) names, which are language or even region-specific: \emph{Ovis
aries}, for example, has valid English vernaculars including lamb,
sheep, wild sheep, and domestic sheep.

@@

A package with the ability to handle the sources of errors outlined
above, and especially while provide an authoritative classification, can
accelerate the work of consuming large volumes of biodiversity data. For
example, this package was used in the process of developing the
\emph{CLOVER} database (\textbf{Gibb2021DatPro?}) of host-virus
associations, by reconciling the names of viruses and mammals from four
different sources, where all of the issues described above were present.

@@

The \texttt{taxonomicdistance} function (and its in-place equivalent,
\texttt{taxonomicdistance!}, which uses memory-efficient re-allocation
if the user needs to change the distance between taxonomic ranks) uses
the (\textbf{Shimatani2001MeaSpe?}) approach to reconstruct a matrix of
distances based on taxonomy, which can serve as a rough proxy when no
phylogenies are available.

@@

\hypertarget{references}{%
\section{References}\label{references}}

>>>>

\texttt{NCBITaxonomy.jl} is a package designed to facilitate the
reconciliation and cleaning of taxonomic names, using a local copy of
the NCBI taxonomic backbone (Federhen 2012; Schoch \emph{et al.} 2020);
The basic search functions are coupled with quality-of-life functions
including case-insensitive search and custom fuzzy string matching to
facilitate the amount of information that can be extracted automatically
while allowing efficient manual curation and inspection of results.
\texttt{NCBITaxonomy.jl} works with version 1.6 of the Julia programming
language (Bezanson \emph{et al.} 2017), and relies on the Apache Arrow
format to store a local copy of the NCBI raw taxonomy files. The design
of \texttt{NCBITaxonomy.jl} has been inspired by similar efforts, like
the R package \texttt{taxadb} (Norman \emph{et al.} 2020), which
provides an offline alternative to packages like \texttt{taxize}
(Chamberlain \& Szöcs 2013).

@@

Unambiguously identifying species is a far more challenging task than it
may appear. There are a vast number of reasons for this. Different
databases keep different taxonomic ``backbones,'' \emph{i.e.} different
data structures in which names are mapped to species, and organised in a
hierarchy. Not all names are unique identifiers to groups. For example,
\emph{Io} can either refer to a genus of plants from the aster family,
or to a genus of molluscs; the genus \emph{Mus} (of which the house
mouse \emph{Mus musculus} is a species), contains a sub-genus
\emph{also} named \emph{Mus}. Conversely, the same species can have
several names, which are valid synonyms: for example, the domestic cow
\emph{Bos taurus} admits \emph{Bos primigenius taurus} as a valid
synonym. Taxonomic nomenclature also changes regularly, with groups
being split, merged, or moved to a new position in the tree of life;
this is, notably, a common occurrence with viral taxonomy, each
subsequent version of which can differ markedly from the last; compare,
\emph{e.g} Lefkowitz \emph{et al.} (2018) to Walker \emph{et al.}
(2020).

To add to the complexity, one must also consider that most taxa names
are at some point manually typed, which has the potential to introduce
additional mistakes in raw data; it is likely to expect that such
mistakes may arise when attempting to write down the (perfectly valid)
names of the bacterial isolate known as \emph{Myxococcus
llanfairpwllgwyngyllgogerychwyrndrobwllllantysiliogogogochensis}, or of
the crowned slaty flycatcher \emph{Griseotyrannus
aurantioatrocristatus}. These mistakes are more likely when dealing with
hyper-diverse samples, like plant census (Dauncey \emph{et al.} 2016;
Wagner 2016; Conti \emph{et al.} 2021). In addition to binomial names,
the same species can be known by many vernacular (common) names, which
are language or even region-specific: \emph{Ovis aries}, for example,
has valid English vernaculars including lamb, sheep, wild sheep, and
domestic sheep.

@@

A package with the ability to handle the sources of errors outlined
above, and especially while provide an authoritative classification, can
accelerate the work of consuming large volumes of biodiversity data. For
example, this package was used in the process of developing the
\emph{CLOVER} database (Gibb \emph{et al.} 2021) of host-virus
associations, by reconciling the names of viruses and mammals from four
different sources, where all of the issues described above were present.

@@

The \texttt{taxonomicdistance} function (and its in-place equivalent,
\texttt{taxonomicdistance!}, which uses memory-efficient re-allocation
if the user needs to change the distance between taxonomic ranks) uses
the Shimatani (2001) approach to reconstruct a matrix of distances based
on taxonomy, which can serve as a rough proxy when no phylogenies are
available.

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Bezanson2017JulFre}{}%
Bezanson, J., Edelman, A., Karpinski, S. \& Shah, V. (2017). Julia: A
Fresh Approach to Numerical Computing. \emph{SIAM Review}, 59, 65--98.

\leavevmode\hypertarget{ref-Chamberlain2013TaxTax}{}%
Chamberlain, S.A. \& Szöcs, E. (2013). Taxize: Taxonomic search and
retrieval in R. \emph{F1000Research}, 2, 191.

\leavevmode\hypertarget{ref-Conti2021MatAlg}{}%
Conti, M., Nimis, P.L. \& Martellos, S. (2021). Match Algorithms for
Scientific Names in FlorItaly, the Portal to the Flora of Italy.
\emph{Plants}, 10, 974.

\leavevmode\hypertarget{ref-Dauncey2016ComMis}{}%
Dauncey, E.A., Irving, J., Allkin, R. \& Robinson, N. (2016). Common
mistakes when using plant names and how to avoid them. \emph{European
Journal of Integrative Medicine}, 8, 597--601.

\leavevmode\hypertarget{ref-Federhen2012NcbTax}{}%
Federhen, S. (2012). The NCBI taxonomy database. \emph{Nucleic acids
research}, 40, D136--D143.

\leavevmode\hypertarget{ref-Gibb2021DatPro}{}%
Gibb, R., Albery, G.F., Becker, D.J., Brierley, L., Connor, R., Dallas,
T.A., \emph{et al.} (2021). Data proliferation, reconciliation, and
synthesis in viral ecology. \emph{bioRxiv}, 2021.01.14.426572.

\leavevmode\hypertarget{ref-Lefkowitz2018VirTax}{}%
Lefkowitz, E.J., Dempsey, D.M., Hendrickson, R.C., Orton, R.J., Siddell,
S.G. \& Smith, D.B. (2018). Virus taxonomy: The database of the
International Committee on Taxonomy of Viruses (ICTV). \emph{Nucleic
Acids Research}, 46, D708--D717.

\leavevmode\hypertarget{ref-Norman2020TaxHig}{}%
Norman, K.E.A., Chamberlain, S. \& Boettiger, C. (2020). Taxadb: A
high-performance local taxonomic database interface. \emph{Methods in
Ecology and Evolution}, 11, 1153--1159.

\leavevmode\hypertarget{ref-Schoch2020NcbTax}{}%
Schoch, C.L., Ciufo, S., Domrachev, M., Hotton, C.L., Kannan, S.,
Khovanskaya, R., \emph{et al.} (2020). NCBI Taxonomy: A comprehensive
update on curation, resources and tools. \emph{Database}, 2020.

\leavevmode\hypertarget{ref-Shimatani2001MeaSpe}{}%
Shimatani, K. (2001). On the Measurement of Species Diversity
Incorporating Species Differences. \emph{Oikos}, 93, 135--147.

\leavevmode\hypertarget{ref-Wagner2016RevSof}{}%
Wagner, V. (2016). A review of software tools for spell-checking taxon
names in vegetation databases. \emph{Journal of Vegetation Science}, 27,
1323--1327.

\leavevmode\hypertarget{ref-Walker2020ChaVir}{}%
Walker, P.J., Siddell, S.G., Lefkowitz, E.J., Mushegian, A.R.,
Adriaenssens, E.M., Dempsey, D.M., \emph{et al.} (2020). Changes to
virus taxonomy and the Statutes ratified by the International Committee
on Taxonomy of Viruses (2020). \emph{Archives of Virology}, 165,
2737--2748.

\end{CSLReferences}
